Safe Transferable Regions

### 2016-07-03 paper/type-inference.tex
Suggestion for the type inference section.

\section{Type Inference}
\label{sec:type-inference}

\name's region type system imposes a heavy annotation burden, and
manually annotating C\# standard libraries with region types
can be tedious. We now present our region type inference algorithm
that eliminates the need to write region type annotations.
% except on some higher-order functions.
Formally, the type inference
algorithm is an elaboration function from programs in $\absof{\FB}$
(i.e., \FB without region types, but with \C{letregion} and \C{open}
expressions, similar to the language introduced in
\S~\ref{sec:overview}) to programs in \FB.

\paragraph{Overview.}

Fig.~\ref{fig:type-inference-algo} presents the high-level outline of the type
inference algorithm.
The algorithm consists of the following steps:
\begin{enumerate}
 \item \emph{Region Parametrization}.
   The first step elaborates the input program by introducing \emph{formal region parameters}
   (for each class and method), and \emph{region variables} (representing yet undetermined
   \emph{actual region parameters}). We also introduce for each class and method, a
   \emph{predicate variable} ($\varphi$) to denote an undetermined set of constraints
   over the region parameters of that class/method.

%%    as follows:
%% \begin{enumerate}
%% \item For every class determine a set of \emph{region parameters} ($\rho$)
%%        it should be parametrized over.
%% \item For every field of the class, determine its region parametrized type by
%%        generating as many fresh \emph{actual region parameters} as required by its class (type),
%%        which in turn become formal region parameters of the containing class.
%% \item  For every method and function type, determine its corresponding region
%%    type template by introducing fresh region parameters as required by
%%    the type of each parameter and return value (in addition to the allocation region
%%    parameter).
%%  \item Elaborate every \C{new} expression, method call or function application by
%%    introducing \emph{fresh region variables} (as required by the region type
%%    template of the called method or function).
%% \item Introduce for each class and method, a \emph{predicate variable} ($\varphi$) to denote
%%    an undetermined set of constraints over the region parameters of that
%%    class/method.
%% \end{enumerate}

 \item \emph{Constraint Generation}.
   In the second step, we analyze the program to generate a set of constraints
   (over the region identifiers and the predicate variables)
   that must hold (as per the static semantics in Fig.~\ref{fig:fb-staticsem}).

 \item \emph{Constraint Solving}.
   We solve the generated set of constraints using our fixpoint constraint
   solving algorithm \csolvestar, which reduces the constraint solving problem to
   an abduction problem. If the original program in $\absof{\FB}$ contains unsafe
   references, for example, a reference from a transferable region to a
   stack region, then the constraints generated during the elaboration
   are not satisfiable. In such a case, \csolvestar{} fails to solve
   the constraints.
% in a Herbrand constraint system, and then relies on \csolve,
% our abduction solver for that domain.

 \item If the solver succeeds, it returns substitution functions $\substFn_\rho$ and
  $\substFn_\varphi$ for free region and predicate variables, respectively, introduced in
  step 1. We apply these substitutions to the elaborated program to produce the final program.
\end{enumerate}

\begin{figure}
\begin{numcodeml}
Infer ($p$) =
  let $q$ = IntroduceRegionParameters($p$) in
  let $(r,C)$ = GenerateConstraints($q$) in
  match (SolveConstraints($C$) with
  | None $\longrightarrow$ None
  | Some ($s$) $\longrightarrow$ Some (ApplySolution $(s,r)$)
\end{numcodeml}

\caption{The type inference algorithm}
\label{fig:type-inference-algo}
\end{figure}

We later on establish the soundness of the type inference algorithm, and also establish
partial completeness results.

\subsection{Region Parametrization for Classes}
\label{sec:fb-templatization}

Region parametrization is an iterative process involving the following three steps,
the first two of which are mutually dependent on each other.

\emph{Introduction of Formal Region Parameters}.
For every class \C{C}, we identify a sequence of formal region parameters
$\pi_0, \cdots \pi_n$ that \C{C} should be parametric over.

\emph{Introduction of Actual Region Parameters}.
We then replace every instance of class \C{C} in the program by an instance
$\C{C}\langle \rho_0, \cdots, \rho_n \rangle$, where $\rho_0, \cdots, \rho_n$
are fresh identifiers denoting actual region parameters.

\emph{Predicate Variable Introduction}. For every class \C{C}, we introduce
a fresh predicate variable $\varphi$, which represents the yet undetermined
outlives constraints between the formal region parameters of class \C{C}.

We identify the region parameters of classes as follows.

\emph{Non-Recursive Classes}.
The class \C{Object} is defined to have a single region parameter $\pi_0$ (the allocation region).
The region parameters for any other non-recursive class \C{C} is determined
only after the region parameters of any class that \C{C} depends on have been
determined: this includes the base-class \C{B} of \C{C} and the class (type)
of any of its fields.
We replace every dependee type \C{T} in \C{C} by its instantiated type,
using fresh region parameters as needed.
The sequence of region parameters for \C{C} is defined to be
the sequence of region parameters for the base class \C{B} concatenated
with the list of all  fresh region parameters introduced while instantiating the types
of the fields in the class.
(The class inherits its allocation region from its base class. Note that if
a class does not specify an explicit base class, it has an implicit base class
\C{Object}.)

This transformation is illustrated below, using a non-generic \C{Pair} class:

\begin{tabular}{ccc}
\begin{minipage}{0.28\linewidth}
\begin{codejava}
class Pair
  $\extends$ $\ObjZ$
{
  $\ObjZ$ fst;
  $\ObjZ$ snd;
}
\end{codejava}
\end{minipage}
&
$\Rightarrow$
&
\begin{minipage}{0.5\linewidth}
\begin{codejava}
class Pair $\langle \rho_0, \rho_1, \rho_2 \; | \; \varphi \rangle$
  $\extends$ $\ObjZ \langle \rho_0 \rangle$
{
  $\ObjZ \langle \rho_1 \rangle$ fst;
  $\ObjZ \langle \rho_2 \rangle$ snd;
}
\end{codejava}
\end{minipage}
\end{tabular}

\emph{Recursive Classes}.
The region parameters for a recursive class is computed in
a similar fashion, with the following difference: any recursive
field is ignored while instantiating region parameters for the fields of
the class, and the region parameters of the recursive class are computed
as before. We then do parameter instantiation for all recursive fields,
such that their region annotations (the actual region parameters) are
exactly the same as the (formal) region parameters of the class.
The following example illustrates this for a non-generic \C{List} class.
The resulting class represents a linked list with spine in the region
$\rho_0$ and data objects in the region $\rho_1$.

\begin{tabular}{ccc}
\begin{minipage}{0.28\linewidth}
\begin{codejava}
class List
  $\extends$ $\ObjZ$
{
  $\ObjZ$ data;
  List next;
}
\end{codejava}
\end{minipage}
&
$\Rightarrow$
&
\begin{minipage}{0.5\linewidth}
\begin{codejava}
class List $\langle \rho_0, \rho_1 \; | \; \varphi \rangle$
  $\extends$ $\ObjZ \langle \rho_0 \rangle$
{
  $\ObjZ \langle \rho_1 \rangle$ data;
  $\C{List} \langle \rho_0, \rho_1 \rangle$ next;
}
\end{codejava}
\end{minipage}
\end{tabular}

The above technique can be extended to mutually recursive classes in a
straightforward manner, by simultaneously parametrizing them (and
then instantiating them).

\emph{Type-Parametric Classes}.
Type parameters of classes are handled as follows.
Consider a type-parametric class \C{C $\langle$ T $\extends$ B $\rangle$}.
The parametric type \C{T} is instantiated using the number of region parameters
that its bound \C{B} has. If no bound is specified for \C{T}, the bound is taken
to be \C{Object}, and \C{T} is instantiated with one region parameter.

\emph{Function Types}.
Since \FB{} is higher-order, fields of function type are allowed. The parameter instantiation step
instantiates function types as follows: the type of every parameter as well as the return value is
instantiated with fresh region identifiers, as described earlier, and finally these region identifiers
are generalized as formal region parameters of the function type.
For example, the function type $\C{List} \rightarrow \C{Object}$ is instantiated as
$\inang{\rho_0, \rho_1, \rho_2} \C{List} \inang{\rho_1,\rho_2} \rightarrow \C{Object}\inang{\rho_0}$.

%% Likewise, given the region-annotated definition of
%% \C{Pair} class from \S\ref{sec:fb-syntax} a region type template for a
%% method with FGJ type $\C{Pair}\inang{\C{A},\C{B}} \rightarrow \C{A}$
%% is \footnote{In our exposition, we assume that classes $\C{A}$ and
%% $\C{B}$ are trivial subclasses of $\ObjZ$ with no fields/methods. Like
%% $\ObjZ$, they accept one region parameter - the allocation region of
%% their objects.}\footnote{We abuse arrow notation to also represent
%% types of methods, but unlike function types, there is no allocation
%% region annotation atop the arrow in a method type.}
%% $\inang{\rhoalloc_0,\rho_1,\rho_2,\rho_3 \,|\, \varphi_0}
%% \C{Pair}\inang{\C{A},\C{B}} \inang{\rho_1,\rho_2,\rho_3} \rightarrow
%% \unitZ$, where $\rhoalloc_0$ and $\rho_{1-3}$ are fresh region
%% variables, and $\varphi_0$ is a fresh predicate variable denoting
%% unknown constraints over $\rhoalloc_0$ and $\rho_{1-3}$.

\subsection{Region Parametrization for Methods}

As the next step, we introduce region parameters for every method.
We do this by instantiating the types of all parameters and the
return value (of the method) using fresh region identifiers (as explained previously),
and then generalizing these region identifiers as formal region parameters
of the method. As for the classes, we also introduce a fresh predicate variable
$\varphi$ for every method.

We then consider every method invocation in the program, and introduce
fresh region variables representing the (yet unknown) actual region
parameters for this particular invocation.
%
We similarly perform instantiation for every constructor invocation
of the form \C{new T($\ldots$)}, by instantiating the type \C{T} as
before, turning it into \C{new T$\langle \rho_0, \cdots, \rho_n \rangle$($\ldots$)},
where $\rho_0, \cdots, \rho_n$ are fresh region variables.

\subsection{The Constraint Satisfaction Problem (CSP)}
\label{sec:fb-constraintsem}

The constraint generation algorithm mimics the static type checker, but accumulates
constraints that must hold for the type checking to succeed.

\paragraph{Syntax of Constraints.}
The constraints are expressed in terms of region identifiers ($\pi$ and $\rho$)
and predicate variables ($\varphi$).

\emph{Region identifiers} are of two kinds: \emph{region constants} and \emph{region variables}.
A \emph{region variable} is introduced to represent an unknown actual region parameter
of a method invocation or object alloation, which will be bound to a region constant by the end of the
type inference.
A region constant may be 
(a) a \emph{formal region parameter} of a class or method, or
(b) a \emph{static region identifier} introduced by a \C{letregion} construct, or
(c) an  \emph{open transferable region identifier} introduced by an \C{open} construct.

Every predicate variable $\varphi$ denotes an unknown \emph{region-constraint},
%% over a set of fixed formal region parameters,
where the set of region-constraints $\phi$ is defined by:
\begin{smathpar}
\begin{array}{lcl}
\phi & \coloneqq & true \ALT \rho \outlives \rho \ALT \rho = \rho \ALT \phi \conj \phi \\
\end{array}
\end{smathpar}
We will use the term \emph{validity constraint} to denote an entailment constraint
of the form $\isvalid{\varphi}{\rho_1 \outlives \rho_2}$.

Our constraints also make uses of \emph{pending substitutions} $F$:
% \footnote{we borrowed this terminology from~\cite{ltpldi08}}
A pending substitution serves to bind formal region parameters in $\varphi$ to the actual region parameters
used in a particular context:
\begin{smathpar}
\begin{array}{lcl}
F & \coloneqq & \cdot \ALT [\rho/\rho]F \\
\end{array}
\end{smathpar}
E.g., in the validity constraint $\isvalid{\rgn_1 \outlives
\rgn_2}{[\rgn_1/\rho_1][\rgn_2/\rho_2]\varphi}$, the pending substitution
is $[\rgn_1/\rho_1][\rgn_2/\rho_2]$. Any concrete formula (call it
$\phisol$) over variables $\rho_1$ and $\rho_2$ is a solution to
$\varphi$ if and only if the formula obtained by substituting $\rgn_1$
and $\rgn_2$ for $\rho_1$ and $\rho_2$ (resp.) in $\phisol$ is
deducible from $\rgn_1 \outlives \rgn_2$.

The constraints generated are of the following kinds:
\begin{itemize}

\item Well-formedness constraints of form $\rho \in \rhoenv$,
restricting the domain of unification for a region variable ($\rho$)
to a constant set $\rhoenv = \{ \pi_1, \cdots, \pi_n \}$ of regions in scope,

\item Well-formedness constraints of form $\tywf{\rhoenv}{\varphi}$, restricting the domain of a predicate
variable ($\varphi$) to the set of all possible constraint formulas over a fixed set of 
regions ($\rhoenv = \{ \pi_1, \cdots, \pi_n \}$) in scope, and

\item Validity constraints of the form $\isvalid{\varphi_{\C{C}}}{\varphi_{\C{B}}}$
  (e.g., to indicate that the region-constraint of a derived class \C{C} is stricter than
  the region-constraint of its base class \C{B}).

\item Validity constraints of the form $\isvalid{\varphi_{\C{C}}}{{F(\varphi_{\C{T}})}}$
  (e.g., to indicate the region-constraint of a class \C{C} must imply the
  region-constraint of each of its fields).

\item Validity constraints of form $\isvalid{\varphi_i \conj \phictxt} {\phicstr}$
where $\varphi_i$ is a predicate variable (representing the precondition of a
method to be determined), $\phicstr$ is a region constraint that is \emph{required}
to hold at a particular program point (within the method), and $\phictxt$ is
a region constraint that is \emph{known} to hold at that program point.

%
%% Formulas $\phictxt$ and $\phicstr$ are concrete, i.e., free
%% of predicate variables and pending substitutions. While $\phictxt$
%% captures relationships that are \emph{known} to hold between concrete
%% region identifiers (i.e., $\rgn$'s) when the constraint was generated,
%% $\phicstr$ captures relationships that are \emph{required} to hold
%% among region varibles (i.e., $\rho$'s), or relationships between
%% region variables and identifiers.
%

\item Validity constraints of the form $\isvalid{\varphi_i \conj \phictxt} {F_j(\varphi_j)}$
generated by an invocation of a method with precondition $\varphi_j$ (where $\phictxt$ and
$\varphi_i$ are as above).

\end{itemize}

\input{fb-constraint-generation}

\paragraph{Constraint Generation.}
The constraint generation algorithm is a direct adaption of the type checker:
each type checking rule is modified to produce a set of constraints that must hold
for the type checker to succeed.
%
Fig.~\ref{fig:constraint-gen} illustrates this for the object allocator \C{new} construct.
%
The appendix presents a more detailed algorithm that performs the elaboration and
constraint generation in a single integrated pass.

\paragraph{Constraint Solution}
Solving the set ($C$) of constraints entails finding an
assignment for each predicate variable $(\varphi$) and each region
variable ($\rho$) that occurs free in $C$, that satisfies all
the validity constraints as well as the well-formedness constraints on
$\varphi$ and $\rhobar$.
% To simplify presentation, we think of $C$ as being parameterized on
% $\varphi$ and $\rhobar$, and write it as $C\lbrack \varphi,\rhobar
% \rbrack$. We now formalize the constraint satisfaction problem, and
% its solution.

\begin{definition}
\emph{(\textbf{Constraint Satisfaction Problem (CSP)})}
A constraint satisfaction problem is a tuple
$(\regionConstants, \regionVars, \predVars, \regionDeltaMap, \predDeltaMap, \constraintSet)$,
where $\regionConstants$ is a set of region constants,
$\regionVars$ is a set of region variables,
$\predVars$ is a set of predicate variables,
$\regionDeltaMap : \regionVars \rightarrow 2^{\regionConstants}$ is a function that
specifies a unification domain for each region variable,
$\predDeltaMap : \predVars \rightarrow 2^{\regionConstants}$ is a function that
specifies the domain for each predicate variable,
and $\constraintSet$ is a set of validity constraints in one of the following forms:
\begin{center}
\(
    \isvalid{\phictxt \conj \varphi}{\phicstr}\qquad
    \isvalid{\phictxt \conj \varphi}{F(\varphi)}
\)
\end{center}
The solution to the constraint satisfaction problem is a pair $(\regionSubstFn,\predSubstFn)$,
where $\regionSubstFn$ is a map from $\regionVars$ to $\regionConstants$
and $\predSubstFn$ is a map from $\predVars$ to a region-constraint formula such that
\begin{itemize}
  \item $\regionSubstFn(\rho) \in \regionDeltaMap(\rho)$, for every $\rho \in \regionVars$,

  \item $\predSubstFn$ is well-formed under $\predDeltaMap$
    (i.e., $\tywf{\predDeltaMap(\varphi)}{\predSubstFn(\varphi)}$, for every $\varphi \in \predVars$).

  \item Every sequent in $C$ is valid after substitutions $\predSubstFn$ and $
    \regionSubstFn$.
  %% $C\lbrack (\regionSubstFn,\predSubstFn) \rbrack$ is valid.
\end{itemize}
\end{definition}

\paragraph{Properties of generated constraints}
In the sequel, we will restrict our attention to a special class of CSPs that satisfy
two interesting properties that allow us to solve the constraints efficiently.
The constraints generated our type inference algorithm satisfy these two properties.

The first property is that every region variable is guaranteed to be unified with some
region constant by the constraints\footnote{
There is one special case where this does not hold: a recursive function
that calls itself in a non-terminating fashion. Such a function never
returns a value, and so the return-value can be typed as anything.
We assume that the return value of such a function is typed to be \C{unit},
which will not introduce any region variables at the call-site of such a function.}
This follows because $\FB$ does not admit uninitialized variables.
% , and the type system uniquely determines the region parameters of every value

>>>>

\vspace*{-0.1in}
\section{Type Inference}
\label{sec:type-inference}

\name's region type system imposes a heavy annotation burden, and the
cost of manually annotating C\# standard libraries with region types
is prohibitive. We now present our region type inference algorithm
that completely eliminates the need to write region type annotations,
except on some higher-order functions. Formally, the type inference
algorithm is an elaboration function from programs in $\absof{\FB}$
(i.e., \FB without region types, but with \C{letregion} and \C{open}
expressions, similar to the language introduced in
\S~\ref{sec:overview}) to programs in \FB. The key aspects of this
elaboration are described below.

% We formalize region type inference as an elaboration function from
% programs in $\absof{\FB}$ (i.e., region annotation-erased \FB) to
% programs in \FB.  The elaboration proceeds in four steps. In the first
% step we make use of the observation that region types are
% \emph{refinements} of FGJ types with region annotations and
% constraints over such region annotations, and compute polymorphic
% region type templates for methods and classes from their FGJ types.
% The templates contain region variables ($\rho$) to denote unknown
% region annotations, and predicate variables ($\varphi$) to denote
% unknown constraints over such region annotations. Free region
% variables are generalized in types (hence, polymorphic). Second, we
% elaborate expressions by introducing region variables to denote
% unknown region arguments in \C{new} expressions, method calls and
% function applications. While elaborating expressions, we make use of
% the computed region type templates to build a system of constraints
% that capture well-formedness requirements and subtyping relationships
% between type templates that must hold (as per the static semantics in
% Fig.~\ref{fb-staticsem}) for the elaboration to be valid.  Third, we
% lift expression elaboration and constraint generation to methods and
% classes. Finally, we solve the constraints to compute solutions to
% region ($\rho$) and predicate ($\varphi$) variables.
%
% Due to the presence of region-polymorphic
% recursion in \FB, constraints generated by the algorithm can be
% circular. More precisely, constraints generated can assume the form
% $\varphi \Leftrightarrow \phi \wedge F(\varphi)$, where $F$ is a
% non-idempotent substitution function for region variables in
% $\varphi$. In the constraint solving phase, the algorithm then relies
% on a fixpoint constraint solving algorithm called \csolvestar to solve
% the constraints and determine assignments for unknown region and
% predicate variables.

\vspace*{-0.1in}
\subsection{Region Type Templates}
\label{sec:fb-templatization}

Region type templates are FGJ types extended with fresh region
variables ($\rho$) and predicate variables ($\varphi$) to denote
unknown region annotations and region constraints, respectively. For
instance, if a variable $x$ has type $\ObjZ$ in FGJ, its region type
template is of form $\ObjZ\inang{\rho_0}$, where $\rho_0$ is a fresh
region variable. FGJ type $\RgnZ\inang{T}$ is special; it is always
templatized as $\RgnZ{\inang{T}}\inang{\toprgn}$. Given the
region-annotated definition of \C{Pair} class from
\S\ref{sec:fb-syntax} a region type template for a method with FGJ
type $\C{Pair}\inang{\C{A},\C{B}} \rightarrow \C{A}$ is \footnote{In
our exposition, we assume that classes $\C{A}$ and $\C{B}$ are trivial
subclasses of $\ObjZ$ with no fields/methods. Like $\ObjZ$, they
accept one region parameter - the allocation region of their
objects.}\footnote{We abuse arrow notation to also represent types of
methods, but unlike function types, there is no allocation region
annotation atop the arrow.}
$\inang{\rhoalloc_0,\rho_{1-4} \,|\, \varphi_0}
\C{Pair}\inang{\C{A},\C{B}} \inang{\rho_{1-3}} \rightarrow
\C{A}\inang{\rho_4}$, where $\rhoalloc_0$ and $\rho_{1-3}$ are fresh
region variables, and $\varphi_0$ is a fresh predicate variable
denoting unknown constraints over $\rhoalloc_0$ and $\rho_{1-3}$.
Region type template of a type variable $\C{a}$ is $\C{a}@\rho$, where
$\rho$ is fresh. For a class, a template is computed based on the 
type templates of its fields, constructors and methods. Additional
region and predicate variables are added as necessary. Following template 
of the \C{Pair} class illustrates this point:
% For a class, a template is computed in two steps. In
% the first step, we templatize the types of all its fields, constructor
% arguments, and arguments and return values of all its methods, along
% with the type of its superclass. In the second step, we generalize all
% the free region variables occuring in the templates computed in the
% first step as region parameters of the class. Finally, we add a fresh
% allocation region parameter ($\rhoalloc$) to the list of parameters,
% and introduce a new predicate variable ($\varphi$) to denote unknown
% constraints on region parameters. For example, consider the standard
% FGJ definition of $\C{Pair}\inang{\C{a},\C{b}}$ class, where $\C{a}$
% and $\C{b}$ are the types of $\C{fst}$ and $\C{snd}$ fields,
% respectively. It can be templatized as following :
\vspace*{-0.1in}
\begin{center}
\begin{codejava}
class Pair$\inang{a\extends\ObjZ, b\extends\ObjZ}$
          $\inang{\rhoalloc_0,\rho_{0-4}\,|\,\varphi_0} \extends \ObjZ\inang{\rho_4}$ {
  $a@\rho_0$ fst; $b@\rho_1$ snd;
  Pair($a@\rho_2$ fst, $b@\rho_3$ snd) { 
    super(); this.fst=fst; this.snd=snd; 
  }
  a@$\rho_5$ getFst$\inang{\rhoalloc_1,\rho_5\,|\,\varphi_1}$() { return this.fst; }
}
\end{codejava}
\end{center}
\vspace*{-0.1in}
% \footnote{In reality, the generalized region variables only come
% from the type templates of class's fields and superclass. Method
% types do not contain any free region variables.}
% The template elides $\C{Pair}$'s methods, whose region type
% templates contain no free variables. Among the region parameters of
% the class template, $\rho_{0-4}$ are obtained by generalizing free
% region variables in the types of its class's fields, constructor
% arguments, and its superclass type.  The remaining parameter
% ($\rhoalloc_0$) is a fresh region variable denoting the allocation
% region argument. Fresh predicate variable $\varphi_0$ denotes
% unknown constraints over $\rhoalloc_0$ and $\rho_{0-4}$ that need to
% hold for template to be a well-formed region-annotated class
% definition in $\FB$. 
%
% The type template for a recursively defined class is computed in two
% steps. First, all recursive occurances of the class among the types of
% its fields are ignored and the class is templatized as if it is a
% non-recursive class. Next, all the recursive occurances are
% templatized with respect to the class template computed in the first
% step, such that their region annotations are \emph{exactly} same as
% the region parameters of the class. Applying this technique on a
% recursively defined $\C{LinkedListNode}\inang{\C{a}}$ class, for
% example, results in a template that uses same region variable to
% denote allocation regions of the current node, and the node that the
% current node refers to. 
%
% For example, consider a generic $\C{ListNode}\inang{\C{a}}$ class
% containing two fields: $\C{data}$ of type $\C{a}$ and $\C{next}$ of
% type $\C{ListNode}\inang{\C{a}}$.  where $\C{a}$ is the type of the
% data stored in the linked list node.  To templatize the
% $\C{ListNode}\inang{a}$ class, we first ignore its recursive
% occurance in the type of $\C{next}$ field, and templatize the type
% $\C{a}$ of $\C{data}$ field as $\C{a}@\rho_0$. Based on this type
% template of $\C{data}$ field, we compute the class's template as
% following: \begin{center} $\C{ListNode}\inang{\C{a}\extends\ObjZ}
% \inang{\rhoalloc_0,\rho_0\,|\,\varphi_0} \extends \ObjZ \{...\}$
% \end{center} Next, we trivially templatize the type of $\C{next}$
% field in the body of the class as
% $\C{ListNode}\inang{\C{a}}\inang{\rhoalloc_0,\rho_0}$.  The
% resulting class represents a linked list with spine in the region
% $\rhoalloc_0$ and data objects in the region $\rho_0$.
% Observe that our strategy of trivially templatizing the recursive
% occurances of a class type in its definition ensures that no new
% region variables are introduced.
The type template for for a recursively defined class is computed in
two steps. First, a template is computed by ignoring the recursive
occurrences. Next, all the recursive occurrences are templatized based
on the template computed in first step.  This approach can be extended
to mutually recursive definitions in a straightforward manner. Using
the techniques outlined above, we compute region type templates for
all classes bound in the class table of the $\absof{\FB}$ program
before we proceed to elaborate expressions.

%\input{fb-elabexpr}
%\input{fb-elabmeth}

\subsection{Elaboration and Constraint Generation}

Elaborating $\absof{\FB}$ expressions to $\FB$ expressions involves
(a). replacing core types occurring in expressions with fresh region
type templates, and (b). explicitly instantiating region parameters
with fresh region variables in method calls and function applications.
% This elaboration is performed with respect to the polymorphic type
% templates of classes and methods computed as per
% \S\ref{sec:fb-templatization}. 
Our expression elaboration function\footnote{Supplement contains the
pseudocode for all the elaboration functions.} traverses expressions
in a syntax-directed manner of a type checker, introducing fresh
region type templates for unknown region types, while generating
constraints over region and predicate variables. The precise nature of
generated constraints is explained in \S\ref{sec:constraints}, but in
summary, they capture the relationships between the type templates of
various subexpressions and their well-formedness.

% Function $\elabExpr$, shown in Fig.~\ref{fig:fb-elabexpr}, performs
% this elaboration for a subset of expressions in $\absof{\FB}$
% %, whose corresponding $\FB$ expressions have been ascribed static
% %semantics in Fig.~\ref{fig:fb-staticsem}. 
% $\elabExpr$ is defined under the same context as the expression typing
% judgment in Fig.~\ref{fig:fb-staticsem} with symbols $\A$,$\ralloc$,
% and $\env$ retaining their meaning. The function traverses expressions
% in a syntax-directed manner of a type checker, introducing fresh
% region type templates for unknown region types, while generating
% constraints over region and predicate variables. The precise nature of
% generated constraints is explained in \S\ref{sec:fb-constraintsem},
% but in summary, they capture the relationships between the type
% templates of various subexpressions and their well-formedness. Note
% that $\elabExpr$ returns the region type template of the
% subexpression, which is used to generate constraints for the
% expression. Functions $\typeOk$ and $\subtypeOk$ (definitions not
% shown) used by $\elabExpr$ implement type well-formedness and subtype
% judgments from Fig.~\ref{fig:fb-staticsem}, respectively.

Expression elaboration can be trivially lifted to the level of a
method, and method elaboration to the level of a class, while
accumulating constraints. Besides such constraints, elaboration
functions for methods and classes also generate new constraints by
checking well-formedness and subtype relationships among the types of
their constituents. For instance, method elaboration generates a new
constraint by typechecking the method body against its return type.
% subtype\footnote{Subtyping is type equality in \FB. In \name,
% subtyping between region types is a straightforward extension of
% subtyping between core types.} of method's return type.  
Likewise, the elaboration function for a class generates a fresh
constraint that captures a couple of safety conditions: first, the
allocation regions of objects referred by the instance variables
should outlive the allocation region of the instance itself, and
second, the allocation regions of a class type and its superclass type
must be the same.

Elaborating a class table ($CT$) entails elaborating every class
definition in $CT$, while collecting constraints. The constraints thus
collected are finally solved to obtain assignments for free region and
predicate variables introduced during templatization and elaboration
stages. The solutions are applied to the class table (and to the
artifacts that make up the class table, recursively) to compute a
class table that maps classes to their fully region-annotated
definitions in $\FB$.

% \subsection{Method and Class Elaboration}

% Elaborating a method involves performingexpression elaboration over
% its body, while collecting constraints. Method elaboration can be
% lifted to the level of classes

% Functions $\elabMeth$ and $\elabClass$ shown in
% Fig.~\ref{fig:fb-elabmeth} lift expression elaboration to method and
% class definitions, respectively. Both functions first build a context
% ($\A$) containing a set ($ \rhoenv$) of regions that are currently
% live, a map ($\aenv$) mapping type variables to their bounds, and a
% constraint formula ($\phicx$) capturing constraints over live region
% variables. We use predicate variables ($\varphi$ and $\varphi_m$) to
% capture constraints over variables in $\rhoenv$ denoting the fact that
% such constraints are yet to be inferred. Both elaboration functions
% generate new constraints by checking well-formedness and subtype
% relationships among types of their constituents. A fresh constraint
% ($C_4$) generated by $\elabClass$ captures a couple of safety
% conditions: first, the allocation regions of objects referred by the
% instance variables should outlive the allocation region of the
% instance itself, and second, the allocation regions of a class type
% and its superclass type must be the same.

% Function $\elabMeth$ elaborates a method definition of class $B$. It
% calls $\elabExpr$ with the context $\A$, its allocation context
% parameter ($\rhoallocm$), and a type environment ($\env$) that
% contains region type bindings for all the arguments of the method,
% including the implicit $\C{this}$ argument. The region type template
% returned by $\elabExpr$ for the method body is checked against its
% expected type (derived from the type template of the method)
% generating more constraints. The function then returns the elaborated
% method definition and the set of constraints.

% $\elabClass$ elaborates the definition of a class $B$. It relies on
% $\elabCons$ (not shown) and $\elabMeth$ functions to elaborate $B$'s
% constructor ($k$) and method definitions ($\bar{d}$), respectively. To
% the set of constraints returned by these functions, $\elabClass$ adds
% constraints generated by checking the well-formedness of the type
% templates of its superclass and fields, and also a new constraint
% capturing a couple of safety conditions: first, the allocation regions
% of objects referred by the instance variables should outlive the
% allocation region of the instance itself, and second, the allocation
% regions of a class type and its superclass type must be the same.

% Function $\elabClassTable$ (not shown) elaborates
% every definition in the class table $CT$, while accumulating
% constraints. The constraints are finally solved for free region and
% predicate variables introduced during templatization and elaboration
% stages. The solutions are applied to the class table (and to the
% artifacts that make up the class table, recursively) to compute a
% class table that maps classes to their fully region-annotated
% definitions in $\FB$.

% Note that if the original program in $\absof{\FB}$ contains unsafe
% references, for example, a reference from a transferable region to a
% stack regions, then the constraints generated during the elaboration
% are not satisfiable. In such case, $\solve$ fails to solve
% constraints, causing the program to be rejected.

\subsection{Constraints}
\label{sec:constraints}

Three kinds of constraints are generated during elaboration:
\begin{itemize}
\item Well-formedness constraints of form $\rho \in \rhoenv$,
restricting the domain of unification for a region variable ($\rho$)
to the set ($\rhoenv$) of regions in scope,
\item Well-formedness constraints of form
$\tywf{\rhoenv}{\varphi}$, restricting the domain of a predicate
variable ($\varphi$) to the set of all possible constraint formulas
over region variables ($\rhoenv$) in scope, and
\item Validity constraints of form $\isvalid{\phicx}{\phi}$, where
$\phicx$ and $\phi$ are region constraints (Fig.~\ref{fig:fb-syntax})
extended with predicate variables and \emph{pending
substitutions}\footnote{we borrowed this terminology
from~\cite{ltpldi08}}:
\vspace*{-0.05in}
\begin{smathpar}
\begin{array}{lcl}
\phicx,\phi & \coloneqq & true \ALT \rho \outlives \rho \ALT \rho = \rho 
    \ALT F(\varphi) \ALT \phi \conj \phi\\
F & \coloneqq & \cdot \ALT [\rho/\rho]F \\
\end{array}
\end{smathpar}
\vspace*{-0.1in}
\end{itemize}
A pending substitution ($F$) is a substitution function over region
variables/identifiers. They represent the substitutions that need to
be carried out when a predicate variable ($\varphi$) is replaced by a
concrete formula in a validity constraint. For instance, in the
validity constraint $\isvalid{\rgn_1 \outlives
\rgn_2}{[\rgn_1/\rho_1][\rgn_2/\rho_2]\varphi}$, pending substitution
is $[\rgn_1/\rho_1][\rgn_2/\rho_2]$. Any concrete formula (call it
$\phisol$) over variables $\rho_1$ and $\rho_2$ is a solution to
$\varphi$ if and only the formula obtained by substituting $\rgn_1$
and $\rgn_2$ for $\rho_1$ and $\rho_2$ (resp.) in $\phisol$ is
deducible from $\rgn_1 \outlives \rgn_2$.
% Validity constraints primarily result from checking if a region type
% template of form $B\inang{\tbar}\inang{\rhoalloc,\rhobar}$ is
% well-formed (i.e., $\typeOk$). The corresponding well-formedness rule
% (Fig.~\ref{fig:fb-staticsem}) requires that the instantiation of $B$'s
% formal region parameters with actual region arguments satisfy its
% precondition, where actuals are subsituted for formals. Since $B$'s
% precondition could be unknown, the requirement results in a in a
% validity constraint with predicate variables and pending
% substitutions. Other sources of validity constraints are calls to
% region-polymorphic methods, where similar situation arises. 
% In general, validity constraints generated by our algorithm assume one
% of the following two forms:
% \begin{center}
% \( \isvalid{\phictxt \conj \bigwedge_{i}\varphi_i}
%            {\phicstr} \qquad
%    \isvalid{\phictxt \conj \bigwedge_{i}\varphi_i}
%            {F_j(\varphi_j)}
% \)
% \end{center}
% Where $\varphi_i$'s denote the unknown preconditions of the class and
% the method under elaboration. If the constraint is generated while
% checking the well-formedness of a type or elaborating an expression,
% then $\varphi_j$'s denote the unknown preconditions of classes and
% methods that were used in that type or expression. Each use of a
% (region-polymorphic) class or a method has to instantiate region
% parameters, resulting in pending substitutions ($F_j$). Formulas
% $\phictxt$ and $\phicstr$ are concrete, i.e., free of predicate
% variables and pending substitutions. While $\phictxt$ captures
% relationships that are \emph{known} to hold between concrete region
% identifiers (i.e., $\rgn$'s) when the constraint was generated,
% $\phicstr$ captures relationships that are \emph{required} to hold
% among region varibles (i.e., $\rho$'s), or relationships between
% region variables and identifiers. Each region variable occuring in
% $\phicstr$ has an associated well-formedness constraint, which
% specifies its unification domain. The unification domain of a
% constraint is the union of unification domains of all region variables
% occuring in the constraint.
